---
title: Data usage agreements and R
output: 
  rmdformats::html_clean:
    self_contained: TRUE
    highlight: kate
vignette: >
  %\VignetteIndexEntry{Data usage agreements and R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(width = 90)

knitr::opts_chunk$set(collapse = TRUE, comment = NA)

```

The `duawranglr` package is designed with the idea that rather than
setting a new data usage agreement (DUA) for each project in an
ongoing collaboration between researchers and data partners, two
things will happen instead:  

1. A master DUA will be signed that establishes a general framework
   for collaboration alongside multiple levels of data restriction;
   for each new project, these levels (*e.g.*, I, II, & III) can then be
   invoked and used to determine which variables may be shared, with
   whom, and under what conditions.  
2. A associated crosswalk file will list the names of data elements
   that are restricted at each level and be used with this package to
   easily and clearly transform raw restricted data files into those
   that can be shared under the conditions of the master DUA.

This vignette goes through the process of using `duawranglr` functions
to create a sharable data set from a raw administrative file. Though
no two projects are alike, users should generally follow the steps
below.

**Important upfront caveat:** using `duawranglr` does not guarantee data
security. It goes without saying, but users, of course, can simply *not
use* the package when attempting to secure restricted data. What this
package does is offer a framework and a set of useful functions that,
when followed, help users secure data in a clear and replicable
manner.

# Administrative data

After loading some libraries, we'll first read in the raw
administrative data file and take a look.

```{r, eval = FALSE}
library(dplyr)
library(readr)
library(duawranglr)

## read in raw administrative data
df <- read_csv('./admin_data.csv')
df
```

```{r, echo = FALSE}
library(dplyr)
library(readr)
library(duawranglr)

## read in raw administrative data
df <- read_csv('../tests/testthat/testdata/admin_data.csv',
               col_types = cols(sid = col_character(),
                                sname = col_character(),
                                dob = col_character(),
                                gender = col_integer(),
                                raceeth = col_integer(),
                                tid = col_integer(),
                                tname = col_character(),
                                zip = col_integer(),
                                mathscr = col_integer(),
                                readscr = col_integer()
                                )
               )
df
```	

The `admin_data.csv` file contains observations for 9 students and has 10
variables associated with each observation. Of these, 1 uniquely
identifies each student, 6 are associated with the student's personal
characteristics, 2 with each student's teacher, and 2 with the
student's test scores in reading and math. 

Though there is no codebook, it appears that the identifier for each
student, `sid`, may be the student's social security number. As
researchers interested in test scores, we have no need for this highly
protected data element other than for its ability to uniquely identify
a student or allow linking to other records. Since we do not need to
link to other records at the moment, any unique number or string will
work for our purposes. Similarly, we don't really need the student's
last name.

Besides math (`mathscr`) and reading (`readscr`) scores, we may be
interested in some of the other covariates. It's likely that many of
these data elements, however, also carry restrictions of varying
severity. For example, the school may be able to share the student's
race/ethnicity and gender (provided the student is not otherwise
identified) with most approved researchers, but can only share
teachers' names (`tid`) under more tightly restricted scenarios.

This is where our DUA crosswalk file comes in handy.

# Data usage agreement

The crosswalk file, `dua_cw.csv`, for the current DUA looks like this:

|level\_i|level\_ii|level\_iii|
|:------:|:-------:|:--------:|
|sid|sid|sid|
|sname|sname|sname|
|dob|dob||
|gender||
|raceeth||
|tid||
|tname|tname|tname|
|zip|zip|

Each column represents a level, `level_i`, `level_ii`, or `level_iii`, along
with the corresponding data element names that are restricted at that
level. The names are arbitrary as far as the package goes, but in
conjunction with the master DUA, they have meaning:

* **Level I**: The first level produces data sets that can be shared more
  widely, but at the cost of losing access to most data elements in
  the final data set. Only math and reading scores can be shared at
  this level.
* **Level II:** The second level has slightly fewer data element
  restrictions, making it better for more research projects. Data
  produced at this level likely come with more sharing and storage
  restrictions than those produced at the first level.
* **Level III:** The third level has the fewest restrictions: only names
  and the student's id cannot be contained in the final data set. Data
  produced at this level will have the strongest restrictions on who
  can use it an how it is stored.
  
The benefit of this level-plus-crosswalk system is two-fold:

1. Data element restrictions are clearly defined for each level, which
   in turn has it's own clearly defined scope for data storage and
   sharing. When starting a new project under the scope of the master
   DUA, researchers and data partners need only to assign it a proper
   level based on the needs of the analyses.
2. Because the crosswalk is a simple tabular file, data element names
   can easily be added or deleted by data partners who do not
   typically use data analysis software. This helps keep the process
   transparent for all team members.

## Set DUA

After reading in the administrative data, the next step is to set the
DUA crosswalk file. The crosswalk file can be in many different
formats and, in most cases, will be read in automatically no matter
the type. (If using a delimited file that isn't a comma- or
tab-separated value format, give the `delimiter` argument the delimiter
string; if using an Excel file with more than one sheet, give the
`sheet` argument the sheet name or number.) If successful, you will get
message telling you so.

```{r, eval = FALSE}
## set the DUA crosswalk
set_dua_cw('dua_cw.csv')
```
```{r, echo = FALSE}
set_dua_cw('../tests/testthat/testdata/dua_one.csv')
```
## Check DUA options

In case you've forgotten the data elements that are restricted at a
particular level, you can check them using the `see_dua_options()`
function with the `level` argument set to the appropriate level. If you
want to compare restrictions across more than one level, you can give
the `level` argument a vector.

```{r}
## compare level II and III restrictions
see_dua_options(level = c('level_ii', 'level_iii'))
```
Alternately, you can see restrictions at all levels if you leave the
`level` argument at its default `NULL` value.

```{r}
## check all level restrictions
see_dua_options()
```

## Set DUA level

After consultation with our data partner, we've decided that data for
this project need to be set at Level II. Because no level allows us to
use the current unique ID, `sid`, we also need to deidentify the
data. We could just delete the `sid` column, but for reasons discussed
below, it will be better if we use it to make new, non-identifiable
but unique IDs. Therefore, we use additional arguments in
`set_dua_level()` to note that deidentification is required and set the
targeted ID column.

```{r}
## set DUA level
set_dua_level('level_ii', deidentify_required = TRUE, id_column = 'sid')
```

## Check DUA level

As we're preparing the data, we can check our restriction level and
the data element names it restricts using `see_dua_level()`. 

```{r}
## see set DUA level 
see_dua_level(show_restrictions = TRUE)
```

# Deidentify data

So far, we have set the DUA crosswalk as well as the restriction level
we will abide by. But our underlying data has not changed at all:

```{r}
## show administrative data has not changed
df
```

We indicated that the data need to be deidentified, so a good first
step in cleaning the raw data is to convert unique student id, `sid`,
into a similarly unique, but unidentifiable value. 

Why not just generate some random string for each value? Though we
don't care to merge these data with other files, we may need to do so
in the future. If we randomly generate new IDs, discarding the old
ones in the process, we will be stuck.

The `deid_dua()` function does two things:

1. It uses a secure `SHA-2` algorithm to convert sensitive IDs into
   unique hexadecimal strings that are almost impossible back convert;
2. It has the option to save a crosswalk file that links the old
   secure IDs to the new IDs.
   
Clearly, it defeats the purpose of deidentifying IDs if a crosswalk
between old and new travels with the new data. But if the crosswalk
file is keep in a secure location, perhaps on the same server that
hosts the raw administrative data, then old IDs can be retrieved if
necessary by those with the proper clearance to do so.

```{r, echo = FALSE}
## deidentify data
tmpdir <- tempdir()
deid_dua(df, write_crosswalk = TRUE, id_length = 20,
         crosswalk_name = 'tmp', crosswalk_path = tmpdir)
```
```{r, eval = FALSE}
## deidentify data
deid_dua(df, write_crosswalk = TRUE, id_length = 20)
```
Here's what the saved crosswalk looks like:

```{r, echo = FALSE}
## show crosswalk
cw <- read_csv(file.path(tmpdir, 'tmp.csv'), col_types = cols(.default = 'c'))
cw
rm(tmpdir)
```
And here now is the data frame:

```{r}
## show data frame
df
```

# Check data frame
```{r}
## write data to disk with one last check
write_dua_df(df, 'cleaned_data.csv', output_type = 'csv')
```
```{r}
## check
check_dua_restrictions(df)
```
```{r}
df <- df %>% select(-dob)
```
```{r}
## check again
check_dua_restrictions(df)
```

# Write cleaned data frame to disk
```{r, eval = FALSE}
## write data to disk with one last check
write_dua_df(df, 'cleaned_data.csv', output_type = 'csv')
```

# Interactive template
```{r, eval = FALSE}
make_dua_template('clean_data.R')
```
```{r, echo = FALSE}
tmpdir <- tempdir()
make_dua_template('clean_data.R', tmpdir, answer_list = list('N','','N','',''))
writeLines(readLines(file.path(tmpdir,'clean_data.R')))
rm(tmpdir)
```

